---
title: "Deriving normalised mtDNA mutational spectrum for all vertebrates"
author: "Dmitrii Iliushchenko & Alina Mikhailova"
date: "10/7/2024"
output:
      html_document:
      toc: true
      toc_depth: 3
      toc_float: true
      number_section: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r echo=FALSE, message=FALSE}
rm(list=ls(all=TRUE))

library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
```

# Get Neutral ATGC for each species and gene
## Set fourfold deg sites
```{r}
VecOfSynFourFoldDegenerateSites <- c('CTT', 'CTC', 'CTA', 'CTG', 
                                     'GTT', 'GTC', 'GTA', 'GTG', 
                                     'TCT', 'TCC', 'TCA', 'TCG', 
                                     'CCT', 'CCC', 'CCA', 'CCG', 
                                     'ACT', 'ACC', 'ACA', 'ACG', 
                                     'GCT', 'GCC', 'GCA', 'GCG', 
                                     'CGT', 'CGC', 'CGA', 'CGG', 
                                     'GGT', 'GGC', 'GGA', 'GGG')
```

```{r eval=FALSE}
### unzip("../data/POLARIZEDBR_DATA.zip")
List = list.files("../data/POLARIZEDBR_DATA/")

for (i in 1:length(List)){
  #i=1
  infile = paste("../data/POLARIZEDBR_DATA/",as.character(List[i]),sep='') 
  if (length(grep('POLARISED',infile)) > 0){
    Species = gsub('\\..*','',as.character(List[i]))
    Gene = gsub(Species,'',as.character(List[i])); Gene = gsub('.POLARISED.txt','',Gene); Gene = gsub('\\.POLARISED.txt','',Gene); Gene = gsub('\\.','',Gene); 
    GeneSpecies = read.table(infile, header = TRUE)
    GeneSpecies = GeneSpecies[GeneSpecies$BranchPosition == 'External',]
    NumOfSeqSpXG = nrow(GeneSpecies)
    ExternalSeqsTogether = paste(GeneSpecies$MoreShallowNodeSeq,collapse = '')
    ExternalSeqsTogether = unlist(strsplit(ExternalSeqsTogether,'')) # 5700/3
    CodonsVec = c(); StartNuc = 1
    if (length(ExternalSeqsTogether)/3 == round(length(ExternalSeqsTogether)/3)){ # if divide by 3 without the rest
      for (j in 1:(length(ExternalSeqsTogether)/3)){
        CodonsVec = c(CodonsVec,paste(ExternalSeqsTogether[StartNuc : (StartNuc+2)],collapse = ''))
        StartNuc = StartNuc+3}
      AllCodons = length(CodonsVec)        # 1021
      CodonsVecNeutral = CodonsVec[CodonsVec %in% VecOfSynFourFoldDegenerateSites]
      NeutralCodons = length(CodonsVecNeutral) # 1900
      data.frame(table(CodonsVecNeutral))
      
      CodonsVecNeutral = gsub("CTA|GTA|TCA|CCA|ACA|GCA|CGA|GGA",'A',CodonsVecNeutral)
      CodonsVecNeutral = gsub("CTT|GTT|TCT|CCT|ACT|GCT|CGT|GGT",'T',CodonsVecNeutral)
      CodonsVecNeutral = gsub("CTG|GTG|TCG|CCG|ACG|GCG|CGG|GGG",'G',CodonsVecNeutral)
      CodonsVecNeutral = gsub("CTC|GTC|TCC|CCC|ACC|GCC|CGC|GGC",'C',CodonsVecNeutral)
      
      Line=c(Species,Gene,length(CodonsVecNeutral[CodonsVecNeutral == 'A']),
             length(CodonsVecNeutral[CodonsVecNeutral == 'T']),
             length(CodonsVecNeutral[CodonsVecNeutral == 'G']),
             length(CodonsVecNeutral[CodonsVecNeutral == 'C']),
             AllCodons, NeutralCodons, NumOfSeqSpXG)
      if (i == 1) {Final = Line}
      if (i >  1) {Final = rbind(Final,Line)}
    }
  }
}
Final = as.data.frame(Final); 
names(Final)=c('Species','Gene','A','T','G','C','NumberOfAllCodons',"NumberOfFourFoldDegenCodons",'NumberOfSeq')
write.table(Final, "../data/Neutral_nuc_content_from_pipeline.txt", col.names = T, row.names = F)
```

# Merge with Taxa

## Prepare Taxa before Merging
```{r}
Final = read.table("../data/Neutral_nuc_content_from_pipeline.txt", header = T)
Taxonomy = read.table("../data/Taxonomy.txt", header = T)
Taxonomy$rank = NULL
table(Taxonomy$Taxa)
```

## Merging nuc content with taxa, if no taxa detected still hold it with NA
```{r}
Final = merge(Final, Taxonomy, all.x = TRUE, by='Species')
NeutralATGC = Final
No_taxa = unique(NeutralATGC[is.na(NeutralATGC$Taxa),]$Species)
cat("List of Species with no taxonomy:", paste(No_taxa, collapse = "\n"), sep = "\n")
```
# Read raw mutational spectra
```{r}
MUT = read.table("../data/Mutational_spectra_in_Chordata_ML.txt", header = TRUE)
cat("N of species =", length(unique(MUT$Species)))
```

# Prepare MUT df to future merging
## Filtration
### Take only normal substitutions and filter out species with too high fraction (> 5%) of unnormal substitutions
```{r}
VecOfNormalSubstitutions <- c('A_C','C_A',
                              'A_G','G_A',
                              'C_G','G_C',
                              'C_T','T_C',
                              'G_T','T_G',
                              'T_A','A_T')
SP = data.frame(table(MUT$Species)); names(SP) = c('Species','NumberOfAllSubst')
SPN = data.frame(table(MUT[MUT$Subs %in% VecOfNormalSubstitutions,]$Species)); names(SPN) = c('Species','NumberOfNormalSubst')
SP = merge(SP,SPN); SP$FractionOfNormal = SP$NumberOfNormalSubst/SP$NumberOfAllSubst

summary(SP$FractionOfNormal) # how many to delete? ask to have more than 95% of substitutions as normal
SpeciesToDelete = SP[SP$FractionOfNormal <=0.95,]$Species; cat("Deleted species = ", length(SpeciesToDelete), "\n");
MUT = MUT[!MUT$Species %in% SpeciesToDelete,]
MUT = MUT[MUT$Subs %in% VecOfNormalSubstitutions,]
cat("Final N of species =", length(unique(MUT$Species)))
```

### Filter Only subs occured in 4fold

Here we have mutations before and after filter
```{r}
nrow(MUT)
MUT = MUT[MUT$AncestralAA == MUT$DescendantAA,]
MUT4fold = MUT[MUT$AncestorCodon %in% VecOfSynFourFoldDegenerateSites & MUT$DescendantCodon %in% VecOfSynFourFoldDegenerateSites,]; 
MUT <- NULL
MUT4fold$Number = 1
nrow(MUT4fold)
```

# Counting and normalization

## Count number of different mutations
```{r}
AGG = aggregate(MUT4fold$Number, by = list(MUT4fold$Species, MUT4fold$Gene), FUN = sum)
names(AGG) = c('Species','Gene','NumOfFourFoldMut')
AGG[is.na(AGG)] <- 0
MUT4fold = merge(MUT4fold,AGG, by = c('Species', "Gene"))
MUT4fold = merge(MUT4fold,NeutralATGC[,c(1:6,9, 10)], by = c("Species", "Gene"))
NeutralATGC = NULL
```

## Calculate number of subs in each species and gene
```{r}
mut_count = aggregate(MUT4fold$Number, by=list(MUT4fold$Species, MUT4fold$Gene, MUT4fold$Subs), FUN=sum); 
names(mut_count) = c('Species','Gene','Subs','Freq')
mut_count =
  mut_count %>% 
  pivot_wider(names_from = Subs, values_from = Freq, values_fill = 0)

MUT4fold = merge(MUT4fold,mut_count, by = c('Species', "Gene"))
```

## Normalization 12 subs
Normalizing by divide substitution on number of all substitution occurred in each species and gene
Also transalte TO HEAVY CHAIN

```{r echo = FALSE}

MUT4fold$T.heavy = as.numeric(MUT4fold$A)
MUT4fold$A.heavy = as.numeric(MUT4fold$T)
MUT4fold$C.heavy = as.numeric(MUT4fold$G)
MUT4fold$G.heavy = as.numeric(MUT4fold$C)

```

### Calculating FrX for 4 nucl
```{r}
MUT4fold$FrA.heavy = MUT4fold$A.heavy / (MUT4fold$A.heavy + MUT4fold$T.heavy + MUT4fold$C.heavy + MUT4fold$G.heavy)
MUT4fold$FrT.heavy = MUT4fold$T.heavy / (MUT4fold$A.heavy + MUT4fold$T.heavy + MUT4fold$C.heavy + MUT4fold$G.heavy)
MUT4fold$FrG.heavy = MUT4fold$G.heavy / (MUT4fold$A.heavy + MUT4fold$T.heavy + MUT4fold$C.heavy + MUT4fold$G.heavy)
MUT4fold$FrC.heavy = MUT4fold$C.heavy / (MUT4fold$A.heavy + MUT4fold$T.heavy + MUT4fold$C.heavy + MUT4fold$G.heavy)
```
### Transalte absolute subs to heavy chain

```{r}
MUT4fold$A_G.heavy = MUT4fold$T_C
MUT4fold$A_T.heavy = MUT4fold$T_A
MUT4fold$A_C.heavy = MUT4fold$T_G
MUT4fold$T_G.heavy = MUT4fold$A_C
MUT4fold$T_A.heavy = MUT4fold$A_T
MUT4fold$T_C.heavy = MUT4fold$A_G
MUT4fold$G_T.heavy = MUT4fold$C_A
MUT4fold$G_C.heavy = MUT4fold$C_G
MUT4fold$G_A.heavy = MUT4fold$C_T
MUT4fold$C_G.heavy = MUT4fold$G_C
MUT4fold$C_T.heavy = MUT4fold$G_A
MUT4fold$C_A.heavy = MUT4fold$G_T
```
### Normalization
```{r}
MUT4fold$A_G.SubstRate.heavy = MUT4fold$A_G.heavy/MUT4fold$A.heavy
MUT4fold$A_T.SubstRate.heavy = MUT4fold$A_T.heavy/MUT4fold$A.heavy
MUT4fold$A_C.SubstRate.heavy = MUT4fold$A_C.heavy/MUT4fold$A.heavy
MUT4fold$T_G.SubstRate.heavy = MUT4fold$T_G.heavy/MUT4fold$T.heavy
MUT4fold$T_A.SubstRate.heavy = MUT4fold$T_A.heavy/MUT4fold$T.heavy
MUT4fold$T_C.SubstRate.heavy = MUT4fold$T_C.heavy/MUT4fold$T.heavy
MUT4fold$G_T.SubstRate.heavy = MUT4fold$G_T.heavy/MUT4fold$G.heavy
MUT4fold$G_C.SubstRate.heavy = MUT4fold$G_C.heavy/MUT4fold$G.heavy
MUT4fold$G_A.SubstRate.heavy = MUT4fold$G_A.heavy/MUT4fold$G.heavy
MUT4fold$C_G.SubstRate.heavy = MUT4fold$C_G.heavy/MUT4fold$C.heavy
MUT4fold$C_T.SubstRate.heavy = MUT4fold$C_T.heavy/MUT4fold$C.heavy
MUT4fold$C_A.SubstRate.heavy = MUT4fold$C_A.heavy/MUT4fold$C.heavy
```


### Drop unused colums
```{r}
MUT_final = subset(MUT4fold, select = -c(AncestralSeqName, DescendantSeqName,
                                         Branch, Number, CodonPosition,
                                         PreviousAncCodon,    AncestorCodon,
                                         NextAncCodon,
                                         PreviousDesCodon,    DescendantCodon,
                                         NextDesCodon,    AncestralAA,
                                         DescendantAA,    Subs, T, A, G, C,
                                         A_C,C_A,
                                         A_G,G_A,
                                         C_G,G_C,
                                         C_T,T_C,
                                         G_T,T_G,
                                         T_A,A_T))

MUT_final = MUT_final[!duplicated(MUT_final[c(1,2)]),]
MUT_final %>% 
  head() %>% 
  kable() %>% 
  kable_styling(full_width = F)
```

### Add Temperature and Maturation
```{r}
enviro = read.table('../data/environment_data.txt', header = TRUE)
growth = read.table('../data/Growth_data.txt', header = TRUE)
maturation = read.table('../data/Maturation_data.txt', header = TRUE)
df_list <- list(enviro, growth, maturation, MUT_final)
All_data <- Reduce(function(x, y) merge(x, y, by = "Species", all = TRUE), df_list)
```
### Save table
```{r}
write.table(All_data, file = '../data/VertebratePolymorphisms.MutSpecDataWithEcology.txt', quote = FALSE)
```
