---
title: "Human mtDNA Analysis: FASTQ Alignment"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

## 1. Introduction

This notebook demonstrates practical steps for analyzing mitochondrial DNA (mtDNA) variants from human NGS data—starting from raw FASTQ files, performing alignment, and supporting downstream variant detection.\
We use open-access tools and real data for reproducibility and hands-on experience. See README for details.

## 2. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rsubread)
library(QuasR)
library(GenomicAlignments)
library(VariantAnnotation)
library(ggplot2)
library(data.table)
library(aws.s3)
```

## 3. Download/Specify Inputs

*Download the data using AWS CLI or SRA tools. FASTQ files and reference genome should be present in the working directory.*

```{r download}
save_object(
  object = "phase3/data/HG00096/sequence_read/SRR062634.filt.fastq.gz",
  bucket = "1000genomes",
  file = "SRR062634.filt.fastq.gz"
)
fastq_file <- 'SRR062634.filt.fastq.gz'
ref_fasta <- 'GRCh38.fa'
```

## 4. Alignment

*Indexing is needed only once for each reference genome version.*

```{r alignment, echo=FALSE}
buildindex(basename="GRCh38_index", reference=ref_fasta)
align(index="GRCh38_index", readfile1=fastq_file, output_file="aligned.bam")
```

-   This command aligns your reads and creates a BAM file.
-   The BAM file can be visualized in IGV or summarized in R.

## 5. BAM Summary and Visualization

```{r plot, echo=FALSE}
summary <- qQC("aligned.bam", genome=ref_fasta)
plot(summary$coverage)
```
